---
title: "Relational Databases"
author: "Dr. Lendie Follett"
date: '`r format(Sys.Date())`'
output:
  xaringan::moon_reader:
      nature:
        highlightStyle: github
        highlightLines: true
        countIncrementalSlides: false
      css: [default, hygge-duke, tamu-fonts]
---


<style type="text/css">
  .reveal p {
    text-align: left;
  }
  .reveal ul {
    display: block;
  }
  .reveal ol {
    display: block;
  }  
</style>


## Combining multiple data sources

* Often (usually) the analyses you encounter will require the use of multiple sources of data. 
* For future discussions, we may refer to datasets (SAS) and data frames (R) simply as tables.
* The collection of multiple tables are called \textbf{relational data} because it's the relationships between the data sets that convey the meaningful information needed to complete an analysis.


---

## Introduction to set theory

* It will be helpful to think of the relationships between tables in terms of verbs used in set theory
* A **set** is a collection of elements.
  - $L = \{1,2,3,4\}$: "Set L contains elements $1,2,3$ and $4$"
  - $R = \{1,3,5,6,7\}$: "Set R contains elements $1,3,5,6$, and $7$"
  
--
  
* Why do we talk about set theory in a programming class?

--

  - It is useful to view tables as sets containing observations. 




---
## Introduction to set theory

.pull-left[
```{tikz, fig.cap = " ", fig.ext = 'png',eval=TRUE,echo=FALSE}
\def\firstcircle{(0,0) circle (1.5cm)}
\def\secondcircle{(0:2cm) circle (1.5cm)}

\colorlet{circle edge}{blue!50}
\colorlet{circle area}{blue!20}

\tikzset{filled/.style={fill=circle area, draw=circle edge, thick},
outline/.style={draw=circle edge, thick}}

\begin{tikzpicture}
\begin{scope}
    %\clip \firstcircle;
        \fill[filled] \firstcircle;
    \fill[filled] \secondcircle;
\end{scope}
\draw[outline] \firstcircle node {$L$};
\draw[outline] \secondcircle node {$R$};
\node[anchor=south] at (current bounding box.north) {$L \cup R$};
\end{tikzpicture}
```
]
.pull-right[
* The **union** of L and R is the set that contains the elements that are either members of L or members of R or members of both L and R.
]

---
## Introduction to set theory

.pull-left[
```{tikz, fig.cap = " ", fig.ext = 'png', eval=TRUE, echo=FALSE}
\def\firstcircle{(0,0) circle (1.5cm)}
\def\secondcircle{(0:2cm) circle (1.5cm)}

\colorlet{circle edge}{blue!50}
\colorlet{circle area}{blue!20}

\tikzset{filled/.style={fill=circle area, draw=circle edge, thick},
outline/.style={draw=circle edge, thick}}

\begin{tikzpicture}
\begin{scope}
    \clip \firstcircle;
    \fill[filled] \secondcircle;
\end{scope}
\draw[outline] \firstcircle node {$L$};
\draw[outline] \secondcircle node {$R$};
\node[anchor=south] at (current bounding box.north) {$L \cap R$};
\end{tikzpicture}

```
]
.pull-right[

* The **intersection** of L and R is the set that contains elements that are members of both L and R.

]


---

## Keys

* In terms of tables, *membership* is determined by one or more **key** variables (columns). There are two types of keys. In RDS (Wickham), these are defined as:

  *  A **primary** key is one or more variables which can uniquely identify an observation in a table.
  *  A **foreign** key uniquely identifies an observation in another table. 

*  A relation between a pair of tables is formed by matching **primary keys** and corresponding **foreign keys**.




---
## Key Examples

Identifying .yellow[primary keys] and .blue[foreign keys].


```{r, echo=FALSE, size = "tiny"}
library(kableExtra)

student <- data.frame(
    student_id = c("6422","3730","5792","9078"),
    last = c("Jones", "Johnson", "Johnson", "Ma"),
    club = rep("stat", 4)
) 

gpa <- data.frame(
  student_id =c("9382","6422", "3730","5792"),
  gpa = c( 3.3, 3.7, 2.9, 4.0),
  major = c("Stat","Stat","Math", "Comp Sci")
)

major <- data.frame(
   major = c("Stat","Math", "Comp Sci", "Mgmt"),
   college = c("LAS", "LAS", "LAS", "Business")
)



#kable(list(student, gpa))
kable(student, caption = "data set: club") %>%
  kable_styling(full_width = FALSE, position = "float_left")%>%
    column_spec(1,  background = "yellow")

kable(gpa, caption = "data set: gpa") %>%
  kable_styling(full_width = FALSE, position = "left")%>%
    column_spec(1,  background = "yellow")%>%
    column_spec(3,  background = "blue")

kable(major, caption = "data set: major")%>%
    column_spec(1,  background = "yellow")
```
???
In simple cases (like the one illustrated), a single variable is enough to identify an observation. 

For example, in the stat club data set, the student id variable is sufficient to identify a row of data. Thus it is a primary key. 

Likewise for the gpa dataset. 

major is a primary key for the major dataset. It is a foreign key for the gpa dataset!

---

## Joins

*  Joins combine a pair of tables (say, table L and table R) based matching keys to form a single table (say, X). 
*  The resulting table, X, will contain variables from both tables. The observations present in X depend on the **join type**.
*  There are several types of joins. We will cover:

  *  Inner join
  *  Left join
  *  Right join
  *  Full join

*  The correct type of join will depend on the particular goal.


---


## Inner Join
```{tikz, fig.cap = " ", fig.ext = 'png', eval=TRUE, echo=FALSE}
\def\firstcircle{(0,0) circle (1.5cm)}
\def\secondcircle{(0:2cm) circle (1.5cm)}

\colorlet{circle edge}{blue!50}
\colorlet{circle area}{blue!20}

\tikzset{filled/.style={fill=circle area, draw=circle edge, thick},
outline/.style={draw=circle edge, thick}}

\begin{tikzpicture}
\begin{scope}
    \clip \firstcircle;
    \fill[filled] \secondcircle;
\end{scope}
\draw[outline] \firstcircle node {$L$};
\draw[outline] \secondcircle node {$R$};
\node[anchor=south] at (current bounding box.north) {$L \cap R$};
\end{tikzpicture}
```

* Match observations for all matching pairs of keys.
* Important characteristic: observations corresponding to unmatched keys are **not included in the resulting table**.

???

Set theory analogy: intersection
---
class: middle, center
## Need GPAs of stat club members

---
## Inner Join

```{r, echo=FALSE, size = "tiny"}
library(kableExtra)

student <- data.frame(
    student_id = c("6422","3730","5792","9078"),
    last = c("Jones", "Johnson", "Johnson", "Ma"),
    club = rep("stat", 4)
) 

gpa <- data.frame(
  student_id =c("9382","6422", "3730","5792"),
  gpa = c( 3.3, 3.7, 2.9, 4.0),
  major = c("Stat","Stat","Math", "Comp Sci")
)

major <- data.frame(
   major = c("Stat","Math", "Comp Sci", "Mgmt"),
   college = c("LAS", "LAS", "LAS", "Business")
)



#kable(list(student, gpa))
kable(student, caption = "data set: club") %>%
  kable_styling(full_width = FALSE, position = "float_left")%>%
    row_spec(c(1,2,3),  background = "green")

kable(gpa, caption = "data set: gpa") %>%
  kable_styling(full_width = FALSE, position = "left")%>%
    row_spec(c(2,3,4),  background = "green")

merge(student, gpa, all.x = FALSE, all.y = FALSE) %>%kable( caption = "data set: inner")
```
---



## Left Join
```{tikz, fig.cap = " ", fig.ext = 'png', eval=TRUE, echo=FALSE}
\def\firstcircle{(0,0) circle (1.5cm)}
\def\secondcircle{(0:2cm) circle (1.5cm)}

\colorlet{circle edge}{blue!50}
\colorlet{circle area}{blue!20}

\tikzset{filled/.style={fill=circle area, draw=circle edge, thick},
outline/.style={draw=circle edge, thick}}

\begin{tikzpicture}
\begin{scope}
    %\clip \firstcircle;
    \fill[filled] \firstcircle;
\end{scope}
\draw[outline] \firstcircle node {$L$};
\draw[outline] \secondcircle node {$R$};
\node[anchor=south] at (current bounding box.north) {$L$};
\end{tikzpicture}
```


*  For each observation in $L$, look for matching key in $R$.
*  Keeps all observations in $L$.


---
class: middle, center
## Add GPA information, wherever possible, to the Stat club member table

---
## Left join
```{r, echo=FALSE, size = "tiny"}
library(kableExtra)

student <- data.frame(
    student_id = c("6422","3730","5792","9078"),
    last = c("Jones", "Johnson", "Johnson", "Ma"),
    club = rep("stat", 4)
) 

gpa <- data.frame(
  student_id =c("9382","6422", "3730","5792"),
  gpa = c( 3.3, 3.7, 2.9, 4.0),
  major = c("Stat","Stat","Math", "Comp Sci")
)

major <- data.frame(
   major = c("Stat","Math", "Comp Sci", "Mgmt"),
   college = c("LAS", "LAS", "LAS", "Business")
)

options(knitr.kable.NA = "")

#kable(list(student, gpa))
kable(student, caption = "data set: club") %>%
  kable_styling(full_width = FALSE, position = "float_left")%>%
    row_spec(c(1,2,3,4),  background = "green")

kable(gpa, caption = "data set: gpa") %>%
  kable_styling(full_width = FALSE, position = "left")%>%
    row_spec(c(2,3,4),  background = "green")

merge(student, gpa, all.x = TRUE, all.y = FALSE) %>%kable( caption = "data set: left")
```

---
## Right Join
```{tikz, fig.cap = " ", fig.ext = 'png', eval=TRUE, echo=FALSE}
\def\firstcircle{(0,0) circle (1.5cm)}
\def\secondcircle{(0:2cm) circle (1.5cm)}

\colorlet{circle edge}{blue!50}
\colorlet{circle area}{blue!20}

\tikzset{filled/.style={fill=circle area, draw=circle edge, thick},
outline/.style={draw=circle edge, thick}}
\begin{tikzpicture}
\begin{scope}
    %\clip \firstcircle;
    \fill[filled] \secondcircle;
\end{scope}
\draw[outline] \firstcircle node {$L$};
\draw[outline] \secondcircle node {$R$};
\node[anchor=south] at (current bounding box.north) {$R$};
\end{tikzpicture}
```


* For each observation in $R$, look for matching key in $L$.
* Keeps all observations in $R$.

---

class: middle, center
## Identify whether students in GPA table are stat club members

---
## Right join
```{r, echo=FALSE, size = "tiny"}
library(kableExtra)

student <- data.frame(
    student_id = c("6422","3730","5792","9078"),
    last = c("Jones", "Johnson", "Johnson", "Ma"),
    club = rep("stat", 4)
) 

gpa <- data.frame(
  student_id =c("9382","6422", "3730","5792"),
  gpa = c( 3.3, 3.7, 2.9, 4.0),
  major = c("Stat","Stat","Math", "Comp Sci")
)

major <- data.frame(
   major = c("Stat","Math", "Comp Sci", "Mgmt"),
   college = c("LAS", "LAS", "LAS", "Business")
)

options(knitr.kable.NA = "")

#kable(list(student, gpa))
kable(student, caption = "data set: club") %>%
  kable_styling(full_width = FALSE, position = "float_left")%>%
    row_spec(c(1,2,3),  background = "green")

kable(gpa, caption = "data set: gpa") %>%
  kable_styling(full_width = FALSE, position = "left")%>%
    row_spec(c(1,2,3,4),  background = "green")

merge(student, gpa, all.x = FALSE, all.y = TRUE) %>%kable( caption = "data set: right")
```
---
## Full Join

```{tikz, fig.cap = " ", fig.ext = 'png', eval=TRUE, echo=FALSE}
\def\firstcircle{(0,0) circle (1.5cm)}
\def\secondcircle{(0:2cm) circle (1.5cm)}

\colorlet{circle edge}{blue!50}
\colorlet{circle area}{blue!20}
\tikzset{filled/.style={fill=circle area, draw=circle edge, thick},
outline/.style={draw=circle edge, thick}}
\begin{tikzpicture}
\begin{scope}
    %\clip \firstcircle;
    \fill[filled] \firstcircle;
        \fill[filled] \secondcircle;
\end{scope}
\draw[outline] \firstcircle node {$L$};
\draw[outline] \secondcircle node {$R$};
\node[anchor=south] at (current bounding box.north) {$L \cup R$};
\end{tikzpicture}
```

* Match keys wherever possible.
* Keeps all observations in $L$ and keeps all observations in $R$.
???
set theory analogy: union

---
class: middle, center
## I want to know information about any student who is either in stat club or in the GPA table

---
## Full Join
```{r, echo=FALSE, size = "tiny"}
library(kableExtra)

student <- data.frame(
    student_id = c("6422","3730","5792","9078"),
    last = c("Jones", "Johnson", "Johnson", "Ma"),
    club = rep("stat", 4)
) 

gpa <- data.frame(
  student_id =c("9382","6422", "3730","5792"),
  gpa = c( 3.3, 3.7, 2.9, 4.0),
  major = c("Stat","Stat","Math", "Comp Sci")
)

major <- data.frame(
   major = c("Stat","Math", "Comp Sci", "Mgmt"),
   college = c("LAS", "LAS", "LAS", "Business")
)

options(knitr.kable.NA = "")

#kable(list(student, gpa))
kable(student, caption = "data set: club") %>%
  kable_styling(full_width = FALSE, position = "float_left")%>%
    row_spec(c(1,2,3,4),  background = "green")

kable(gpa, caption = "data set: gpa") %>%
  kable_styling(full_width = FALSE, position = "left")%>%
    row_spec(c(1,2,3,4),  background = "green")

merge(student, gpa, all.x = TRUE, all.y = TRUE) %>%kable( caption = "data set: full")
```
---
## Before we begin...

* Real data is messy.
* If the key being used to merge is a character string, make sure values are spelled consistently across data sets. 
  * e.g., "a7392" and "A7392" will NOT match in a merge/join.
* If key is a character string that involves all numeric characters, make sure leading 0s haven't been lost. 
  * e.g., in Excel, a value of 00073615 will be automatically converted to 73615. If the data was manipulated in Excel before imported to R and SAS, this could present an issue.
  
  
---
## Joins in SAS  
  
* Typically, we create a single SAS data set from another single SAS data set using the `set` statement
within a data step. e.g,

```r
data work.club;
  set s40dat.club;
run;
```

--

* We can replace that with a `merge` statement to accomplish joining two (or more) SAS data sets.

```r
data work.merged;
*  merge s40dat.club s40dat.gpa;
   by student_id;
run;
```
--
* The merge statement will be used in conjunction with a `by` statement that species the common key variable(s).

---
## Joins in SAS


* SAS requires that input data sets be sorted by the common key variable(s) prior to the join.

* Thus, your code sequence may look *something* like:

```r
/*Sort both data sets by key(s)*/
proc sort data = data-x;
by list-key(s);
run;

proc sort data = data-y;
by list-key(s);
run;

/*Perform merge*/
data output-data ;
merge data-x<(in = idxx )> data-y<(in = idxy )>;
by list-key(s) ;
<if expression-involving-idxs >;
run;
```

* Finally, the type of join can be determined using the `in =` data set option (this is new to us) along with an appropriate subsetting if statement (also new to us).

---
## Wait, what was that about `in = `?

* The `merge` statement will ask SAS to perform a full join. Always. 
* We can perform other types of joins by removing certain observations based on membership. 
* The in = data set option creates a temporary binary (0/1)
variable indicating whether the observation was a member of that particular data set.
* Then, we can use those temporary variables (named idxx
and idxy in the below example) to subset and thus perform
the correct join.
* We subset using the `if` statement, which can be used just as we used the `where` statement.
```r
/*Perform merge*/
data output-data ;
merge data-x<(in = idxx )> data-y<(in = idxy )>;
by list-key(s) ;
<if expression-involving-idxs >;
run;
```
---
## Joins in SAS

In each of the following merges, we create temporary, binary variables that we name a and b. Then, we use these to filter (or not!) observations from the full join to create an inner, full, left, or right join.

.pull-left[
```r
/*inner join*/
data merged_data;
merge student(in = a) gpa(in = b);
by student_id;
*if a and b;
run;
```
]
.pull-right[
```r
/*full join*/
data merged_data;
merge student(in = a) gpa(in = b);
by student_id;
*/*NO IF STATEMENT! (keep all observations)*/
run;
```
]

.pull-left[
```r
/*left join*/
data merged_data;
merge student(in = a) gpa(in = b);
by student_id;
*if a;
run;
```
]
.pull-right[
```r
/*right join*/
data merged_data;
merge student(in = a) gpa(in = b);
by student_id;
*if b;
run;
```
]

---
## Case Study

*Have*: A data set containing NFL draft results. A separate data set containing combine results of college football players. 

*Want*: Find meaningful associations between combine results and draft results. Essentially, what athletic measurements are most important for a desirable draft outcome?

---

## Case Study
### Football data: Player stats
A data set containing combine results (various athletic measurements) of college football players from 1999 to 2015. 
```{r, echo=FALSE}

player_stats <- read.csv("Case Studies/Combine Results 1999-2015.csv")

head(player_stats) %>% kable()  %>%
  kable_styling(font_size = 10)
  

```

---
## Case Study
### Football data: Draft results
A data set containing draft results from 1999 to 2015
```{r, echo=FALSE}

draft_results <- read.csv("Case Studies/Draft Results 1999-2015.csv")

head(draft_results) %>% kable()  %>%
  kable_styling(font_size = 10)
```

