
---
title: "SAS Data Sets"
author: "Dr. Lendie Follett"
date: '`r format(Sys.Date())`'
output:
  xaringan::moon_reader:
      nature:
        highlightLines: true
        countIncrementalSlides: false
      css: [ hygge-duke, tamu-fonts]
---
<style type="text/css">
  .reveal p {
    text-align: left;
  }
  .reveal ul {
    display: block;
  }
  .reveal ol {
    display: block;
  }  
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(flair)
library(kableExtra)

colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}

options(width = 999)
```

# Data 

* Accessing data is the first step in any SAS programming project. 
* We will be working with *structured data*, which can easily be read into SAS. 

* Definition: Structured data includes defined <span style="background:blue">rows</span> and columns. You can store structured data in SAS. Oracle, Excel, Hadoop, etc...

<!-- Highlight rows one color, columns another color  -->

```{r, echo=FALSE}

array(".", dim = c(3,4)) %>%
  kbl() %>%
  kable_paper() %>%
  kable_styling() %>%
   row_spec(2, bold = T, color = "white", background = "blue")

```

Another name for a row of data is an *observation*.

---

# Data 

* Accessing data is the first step in any SAS programming project. 
* We will be working with *structured data*, which can easily be read into SAS. 

* Definition: Structured data includes defined rows and <span style="background:green">columns</span>. You can store structured data in SAS. Oracle, Excel, Hadoop, etc...

<!-- Highlight rows one color, columns another color  -->

```{r, echo=FALSE}

array(".", dim = c(3,4)) %>%
  kbl() %>%
  kable_paper() %>%
  kable_styling() %>%
    column_spec(3, bold = T, color = "white", background = "green")

```

Another name for a column is a *variable*.

---

## Anatomy of a SAS data set
* Once data is read into SAS, it is called a *SAS data set.* 
* There are two parts to any SAS data set: 

1. a data portion and
2. a descriptor portion. 

---

## Data Portion

<!-- data values stored in rows and columns (what you imagine when you think "data set") -->
&nbsp;

If you want to see the data portion, you can use a proc print step.
```{r, eval=FALSE}
proc print data = sashelp.iris;
run;
```

```{r, eval=TRUE, echo=FALSE}

iris[1:6,] %>%
  kbl() %>%
  kable_minimal()

```

---
## Descriptor portion:

"Metadata" 
* name of SAS data set
* number of rows
* date created
* column names
* column attributes 

<!-- metadata: name of data set, number of rows, date created, column names, column attributes -->
---
## Descriptor portion
### Column Attributes

Wait a second, what is a "*column attribute*"? 

* This is something all columns in SAS must have. 

* There are three components. 

  1. Name
  2. Type
  3. Length
---

## Descriptor portion
### 1. Name: 


```{r, eval=TRUE, echo=FALSE}

iris[1:4,] %>%
  mutate_all(function(x){x = "."})%>%
  kbl() %>%
  kable_minimal()

```

--

* Must be 1-32 characters, starts with a letter (a-Z,A-Z) or underscore, only contains: (letters, numbers, underscores), can be uppercase or lowercase or mixed

--

* Examples of valid names: stat40, mylib

--

* Examples of invalid names: 40stat, stat.40

???

stat40, mylib,

40stat are invalid names

---

## Descriptor portion
### 2. Type: 

Column type is either (1) numeric or (2) character.

```{r, eval=TRUE, echo=FALSE}

iris[1:6,] %>%
  kbl(align = c("r", "r", "r", "r", "l")) %>%
  kable_minimal()

```

Columns of type numeric contain digits 0-9, minus sign, decimal points, or scientific notation (E). Dates are also understood as numeric! More on this later...

&nbsp;

Character variables can contain letters, numbers, special characters, and blanks. 

---
## Descriptor portion
### 3. Length

* Storage (in bytes) allocated to store column values.
* This is related to the type. 
* Numeric columns, by default, area always stored as 8 bytes (enough for ~ 16 digits). 
* Character can be any length (practically). The length will be set according to how many characters are present in the column and will assume the length of the longest string. 

---
## Descriptor portion
### So, you want to see the descriptor portion?


If you want to see the descriptor portion (which, if you recall, includes the column attributes), you can use a proc contents step. This is a new one!

```{r, eval=FALSE}

proc contents data = sashelp.shoes;
run;

```

---
## Case Study

sashelp.heart contains data from the Framingham heart study. Let's identify the data portion and the descriptor portion, paying special attention to the column attributes. 

```{r, eval=FALSE}

proc print data = sashelp.heart;
run;

proc contents data = sashelp.heart;
run;

```

---
background-color: #6495ED


# Your Turn


In the heart data, how are missing character values represented in the data? What about missing numeric values? 

---

## Where have we been


* We know how a SAS data set is structured
* We know how to learn about the descriptor portion

--

## Where we are going

* SAS libraries


